---
title: 'Reproducible Research: Peer Assessment 1 (Jan. 2016)'
author: "Hongmei Li-Byarlay"
date: "January 9, 2016"
output: html_document
---

Loading and preprocessing the data

```{r}
library(knitr)
opts_chunk$set(echo = TRUE)
library(data.table)
library(ggplot2)
```

#Show any code that is needed to
```{r}
setwd("C:/Users/Hongmei/Desktop/R_coursera/Reproduciable/")
```

#Load the data (i.e. read.csv())
```{r}
rdata <- read.csv(unz("repdata-data-activity.zip", "activity.csv"))
```
#Process/transform the data (if necessary) into a format suitable for your analysis
```{r}
rdata$date <- as.Date(rdata$date, format = "%Y-%m-%d")
rdata$interval <- as.factor(rdata$interval)
names(rdata)
str(rdata)
```



###########################
#What is mean total number of steps taken per day? Let's compute the number of missing vs. complete records
```{r}
originalValue <- complete.cases(rdata)  
nMissing <- length(originalValue[originalValue==FALSE])             # number of records with NA  
nComplete <- length(originalValue[originalValue==TRUE])             # number of complete records
title="Missing vs. Complete Cases"  
barplot(table(originalValue),main=title,xaxt='n', col="gray")       # render Complete Cases barplot  
axis(side=1,at=c(.7,1.9),labels=c("Missing","Complete"),tick=FALSE) # render axis  
text(.7,0,labels=nMissing, pos=3)                                   # label the NA's bar  
text(1.9,0,labels=nComplete, pos=3)
steps_per_day <- aggregate(steps ~ date, rdata, sum)
colnames(steps_per_day) <- c("date","steps")
head(steps_per_day)
```

#ignore the missing values (a valid assumption). We proceed by calculating the total steps per day.
```{r}
steps_per_day <- aggregate(steps ~ date, rdata, sum)
colnames(steps_per_day) <- c("date","steps")
head(steps_per_day)
```

#Now we use a histogram of the total number of steps taken per day, plotted with appropriate bin interval.

```{r}
ggplot(steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "gray", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw()
```

#Now we calculate the mean and median of the number of steps taken per day.

```{r}
steps_mean   <- mean(steps_per_day$steps, na.rm=TRUE)
steps_median <- median(steps_per_day$steps, na.rm=TRUE)
```
#We calculate the aggregation of steps by intervals of 5-minutes and convert the intervals as integers and save them in a data frame called steps_per_interval.
```{r}
steps_per_interval <- aggregate(rdata$steps, 
                                by = list(interval = rdata$interval),
                                FUN=mean, na.rm=TRUE)
```

#convert to integers
##this helps in plotting
```{r}
steps_per_interval$interval <- 
        as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")
```


#################
##We make the plot with the time series of the average number of steps taken (averaged across all days) versus the 5-minute intervals:
```{r}
ggplot(steps_per_interval, aes(x=interval, y=steps)) +   
        geom_line(color="gray", size=1) +  
        labs(title="Average Daily Activity Pattern", x="Interval", y="Number of Steps") +  
        theme_bw()

```

#Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
max_interval <- steps_per_interval[which.max(  
        steps_per_interval$steps),]
```


############
#Imputing missing values
#total number of missing values
```{r}
missing_vals <- sum(is.na(rdata$steps))
missing_vals
```

#Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
#To populate missing values, we choose to replace them with the mean value at the same interval across days. In most of the cases the median is a better centrality measure than mean, but in our case the total median is not much far away from total mean, and probably we can make the mean and median meets.

#We create a function na_fill(data, pervalue) which the data arguement is the rdata data frame and pervalue arguement is the steps_per_interval data frame.
```{r}
na_fill <- function(data, pervalue) {
        na_index <- which(is.na(data$steps))
        na_replace <- unlist(lapply(na_index, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        fill_steps <- data$steps
        fill_steps[na_index] <- na_replace
        fill_steps
}

rdata_fill <- data.frame(  
        steps = na_fill(rdata, steps_per_interval),  
        date = rdata$date,  
        interval = rdata$interval)
str(rdata_fill)
#check if there is NA value remaining or not
sum(is.na(rdata_fill$steps))

###A histogram of the total number of steps taken each day
fill_steps_per_day <- aggregate(steps ~ date, rdata_fill, sum)
colnames(fill_steps_per_day) <- c("date","steps")

##plotting the histogram
ggplot(fill_steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "gray", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day (Count)") + theme_bw()
```

#Are there differences in activity patterns between weekdays and weekends? For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.
#Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
#Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r}
weekdays_steps <- function(data) {
        weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                                    FUN=mean, na.rm=T)
        # convert to integers for plotting
        weekdays_steps$interval <- 
                as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
        colnames(weekdays_steps) <- c("interval", "steps")
        weekdays_steps
}

data_by_weekdays <- function(data) {
        data$weekday <- 
                as.factor(weekdays(data$date)) # weekdays
        weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
        weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))
        
        weekend_steps <- weekdays_steps(weekend_data)
        weekday_steps <- weekdays_steps(weekday_data)
        
        weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
        weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))
        
        data_by_weekdays <- rbind(weekend_steps, weekday_steps)
        data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
        data_by_weekdays
}

data_weekdays <- data_by_weekdays(rdata_fill)
```

```{r}
ggplot(data_weekdays, aes(x=interval, y=steps)) + 
        geom_line(color="darkgray") + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of Steps") +
        theme_bw()
```


